A decentralized network support middleware for MMOGs

### 2009-09-09 trunk/textos/PlanoDoutorado/Lugano/short_planodoc_lugano.tex

% a data deve ser a da defesa; se nao especificada, são gerados
% mes e ano correntes
\date{setembro}{2009}

@@

%
% palavras-chave
% iniciar todas com letras minúsculas, exceto no caso de abreviaturas
%
\keyword{jogos online maciçamente multijogador}
\keyword{MMOG}
\keyword{simulação interativa distribuída}

@@

\begin{abstract}

This doctorate study plan proposes the research and creation of a novel support model for MMOGs (massively multiplayer online games), by using large-scale systems similar to those created for sharing of computing resources, such as file sharing peer-to-peer networks. Following this model, a distributed system could be formed to act as a server for an MMOG. Then, a portion of the game's communication and processing load would be delegated to each node of the server system -- more precisely, the game's virtual environment would be partitioned into regions, each of which managed by one or more server nodes. Such a system would be perceived by the players as an unique central servel to which they connect via a gateway. Upon connected, each player is seamlessly redirected to the appropriate server node. However, for such server system to work as expected, some issues must be addressed, such as maintaining the game state consistent among the different game server nodes, so each player has the same vision; management and balancing of the load assigned to the different server nodes; what to do when a node crashes or leaves the server system and, finally, how to broadcast and store the game state to the players in a distributed manner so that its most recent state can be restored even in the presence of system failures.

\end{abstract}

\chapter{Introduction}

@@

Massively multiplayer online games (MMOGs) are the theme of this study plan. Usually a client-server infrastructure is used to support this kind of games. This plan proposes the investigation and definition of a novel distribution model to be used for MMOGs, reducing the maintenance cost of the game server. In the next sections, the state of the art regarding this area will be presented.

\section{Multiplayer games}

% o que são
% breve histórico e exemplos
% requisitos computacionais (sem mtos detalhes)

In the past few decades, since the creation of the personal computer, computer games have enjoyed an evergrowing popularity. After the acess to the Internet became widely available to the general public, with increasingly fast and cheap domestic connections, multiplayer online games have become an important market. To illustrate this, a study from 2006 estimated that the online games market would grow from US\$ 3.4 billion in 2005 to US\$ 13 billion in 2011 \cite{intelligence2006online}. The main characteristic of the MMOGs is the large number of concurrent players interacting with one another through the Internet. This kind of game may be divided into several categories, such as:

\begin{itemize}

\item FPS: \emph{first-person shooting}, in which each player controls a character who has one or more weapons, with which he fights enemies that, in turn, may be controlled by other players or by artificial intelligence. These games are characterized by fast action, including shooting, moving, dodging etc.;

\item RTS: \emph{real-time strategy}, where each player commands an army, similarly to a chess game, except that the game play is not divided in turns, i.e. it's a real-time game and the player who thinks faster than the others may have an advantage. Nevertheless, the action is usually slower than in an FPS game;

\item RPG: \emph{role-playing game}. In this kind of game, each player has a characters who evolves with time, acquiring power and accumulating wealth. These games are characterized by not having a predetermined end of the ``match'', creating a virtual world with persistent state. The player may, then, connect to this world, play for some time with his character, disconnect and, later, reconnect and continue from the point he was when he disconnected for the last time, for his state is stored in the game server.

\end{itemize}

As examples of well-known online multiplayer games, we may cite \emph{Quake} \cite{quake} as a classic FPS game; \emph{Starcraft: Brood War} \cite{broodwar} a classic RTS game still played nowadays and Diablo II \cite{diablo2} as an RPG example.

\section{Massively multiplayer online games (MMOGs)}
%TODO: por um monte de referências a MMOGs, do tipo (de maior sucesso são X, Y, Z), outros exemplos são A, B, C
% requisitos computacionais elevados a 1000

After single-player computer games evolved to multiplayer online games with the advent of the Internet, MMOGs emerged as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. This kind of games have become very popular in the latest years because they allow the online interaction of a massive number of players at the same time. In the most succesful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. These players can, then, interact with one another and with the virtual environment of the game.

One of the main features of these games, present in almost all commercial MMOGs, is that the matches are very long. In some MMORTS (real-time strategy MMOGs), such as Travian \cite{travian}, the games come to last from six months to one year. MMORPGs (MMOGs in RPG style), as is the case of World of Warcraft, the games have no given time to end. The worlds of these games are so vast and populous -- as regards the number of players -- that they may evolve and modify their state independently of some players being playing it or not, because there is always someone connected so the game never stops. For this reason, it is expected that each player can disconnect and continue playing later, without having to start from scratch. Moreover, it is expected that changes made on the game world persist, no matter how many times each player enters and leaves.

Generally, each player controls an entity called \emph{avatar}, which is nothing more than his representation in the virtual environment -- a character who executes the orders given by the player, therefore interfering in the history and in the outcome of the game. For the game to remain consistent for the different participants, each action performed by each avatar must be processed by the server, which calculates the game state resulting from such action. This new state is then broadcast to other players. Other changes in the game world, such as destruction or creation of objects in the environment, and events such as actions of characters controlled by the server must also be sent to the players involved. It is then through these messages sent by the player to the server and in the opposite direction that players can interact with each other and with the simulated world of the game.

However, it is perceived that such a mechanism can easily saturate the bandwidth and overload the cpu(s) of the server. What is usually done is build an infrastructure with a large central server, usually a cluster with its nodes connected through a local high-speed and low latency network, with an Internet connection of some hundreds or thousands of MBps. The greatest cost comes from maintaining this connection, whose main purpose is to maintain every player as up-to-date as possible about the game state -- naturally, tolerating some delay, which should not be over a given threshold.

%Jogos dessa natureza têm alguns requisitos \cite{schiele2007rpp}, dentre os quais podem ser citados: \textbf{persistência} do estado do mundo virtual e dos avatares; \textbf{disponibilidade} do serviço; \textbf{consistência} do estado do mundo e dos avatares entre os diferentes participantes; \textbf{resistência a trapaça}, no que se refere a tentativas de subversão das regras do jogo por jogadores desonestos e \textbf{escalabilidade}, tendo em vista que o grande número de jogadores poderá facilmente sobrecarregar a infra-estrutura de suporte se esta não tiver capacidade de processamento e, principalmente, largura de banda suficientes.

\chapter{Motivation and state of the art}

%O que é comum a qualquer um dos gêneros de jogos online multijogador é o fato de serem simulações interativas de tempo-real. São simulações no sentido de que simulam um mundo tão parecido quanto possível com o mundo real -- embora alguns jogos se distanciem propositalmente do realismo. A interatividade vem do fato de que há vários participantes executando ações que modificarão o andamento do jogo. Por fim, são de tempo-real porque diversas ações executadas por um jogador devem ter efeito imediato, sob pena de prejudicar alguém. Em um jogo FPS, por exemplo, espera-se que após um jogador enviar o comando de puxar o gatilho, arma de seu personagem dispare o mais rápido possível. Além disso, é necessário manter a consistência da simulação entre os jogadores, para que estes visualizem o jogo no mesmo estado.

%Para jogos online multijogador de pequena escala (com, no máximo, algumas dezenas de participantes), costuma ser suficiente montar uma rede cliente-servidor, onde o servidor é o computador de um dos próprios jogadores, que a ele conectam-se através de um software cliente. Esse servidor atua como árbitro, de forma a manter o estado do jogo consistente entre os diferenetes participantes. Outra abordagem é formar uma pequena rede par-a-par entre os jogadores, utilizando algum algoritmo para manter a consistência da simulação, que agora é distribuída.

As already mentioned, MMOGs are characterized by a large number of players. To support this great number of participants, a client-server infrastructure is generally used, and all interaction is done through the server machine. In this paradigm, the server computer has the official copy of the game state and it is responsible for computing its change over time, while the game client is responsible for presenting to the user the game state sent by the server. The server is also responsible for updating the clients, in real time, on the changes that occur in the virtual environment of the game. Because of the high number of participants, the centralization on the game server makes the maintanance cost very high. Generally, it is required that the MMOG server has a connection to the Internet with several GBps of bandwidth and enough processing power to run the simulation of the virtual world \cite{feng2007wnn}. It becomes impossible, for example, the maintenance of massively multiplayer online games by small or independent groups on a limited budget.

Because of this high cost, alternative solutions have been sought in order to distribute somehow the communication and processing loads originated from the game. Generally, it is proposed the use of peer-to-peer networks -- or even computer GRIDS \cite{butterfly, wang2004gem}. However, an MMOG is a simulation of a dynamic virtual environment, with a maximum tolerable delay between sending the player's actions and receiving the resulting state. MMOGs have also other requirements that must be observed carefully so that the players may have a good game experience. For example, as in a fully decentralized peer-to-peer MMOG the peers may have to calculate themselves the outcome of the players' actions, a malicious player could try to hack the software running in his computer in order to forge game states that provide him unfair advantages over the other players. To distribute the network support structure among different machines connected via the Internet -- which is the case of the peer-to-peer networks -- while meeting the requirements of an MMOG becomes a challenging task that has motivated many researchers in recent years.

In the following sections, we introduce some requirements of the MMOGs in addition to the state of the art as regards the distribution of the infrastructure to support this kind of game. Finally, it will be presented some research questions that are still open, and which can be addressed in a doctoral project.

\section{Requirements of the MMOGs}

Based on the current state of the art and in previous work -- such as in \cite{bezerra2009lbs, bezerra2008a3, vilanova2008psm, cecin2006pps, cecin2004fsa} --, we can summarize a few requirements that we find critical for the game network support infrastructure to be considered adequate, and which will be dealt with in the doctoral course:

\begin{itemize}

	\item \textbf{Persistence} - MMOG games generally consist of a virtual world where players interact with each other and with the environment, changing the state of the world and of their own avatars. It is expected that each player can leave the game and return some time later without these changes having been undone. For this reason, it is necessary to support the persistence of the states of each avatar and of the virtual world as a whole;

	\item \textbf{Availability} - In general, MMOGs require that the players make some kind of payment for account maintenance and, because of that, they tend to be demanding when it comes to availability. Thus, it is desired the system is always available;

	\item \textbf{Consistency} - Multiplayer games consist of a virtual environment shared among the players participating in the same match. It is therefore necessary that the environment is represented in the same way on the machines of the various players so that they can interact with each other and with the world in an appropriate manner;

	\item \textbf{Security} - In a system where there are tens of thousands of participants, it is unreasonable to rely on all of them, requiring some technique to provide resistance against faulty or malicious behaviour. Security here refers not only to the authentication of registered users, but also to check whether the states or actions that the different players submit are valid. Cheaters, for example, can change the software running on their machines so it may send states or actions that give them an advantage in the game;

	\item \textbf{Scalability} - Taking into account the number of participants in an MMOG, it is necessary that the system that supports it is able to maintain the quality of service despite the increasing processing and communication load required for this.

\end{itemize}

\section{Related work}

The \textbf{client-server architecture} meets various aspects necessary for the satisfactory implementation of games such as MMOGs, what includes a high level of control over the system as a whole, which facilitates authentication, persistence and security. But the maintenance of a traditional MMOG server is expensive, as already mentioned, besides making the server a possible bottleneck. In order to minimize this problem some alternatives have been proposed. One of them is to use cluster computing, where a cluster, rather than a single computer, plays the role of the server. This approach has a significant increase in processing power, but does not solve all the problems of massively multiplayer online games. It should also provide the bandwidth needed to support the traffic between the server and the players.

%TODO: citar o paper sigcomm 2007 que mostra os "Challenges in peer-to-peer gaming"

Another approach is the \textbf{peer-to-peer architecture}, where the simulation is divided among the computers involved. There can be a system without any server, where the peers, which are the machines of the players, come into some kind of agreement for the various steps of the simulation, calculating the game state resulting of each action of the participants. Regarding scalability, this approach is not optimal, since ensuring such "agreement"{} is costly in terms of message exchange \cite{lamport1982bgp}. Even if one of the peers is elected to decide the course of the simulation, still persists the problem of all peers potentially needing to exchange messages with one another. Having $n$ peers, there is a complexity of O({$n^2$}) message exchanges for each step of the simulation, in the worst case. Clearly, this approach is not as scalable as should be a system to support a massively multiplayer online game. Moreover, it would be necessary to distribute the storage and retrieval of the states of the game.

Some studies -- such as \cite{schiele2007rpp, rieche2007ppb, hampel2006ppa, elrhalibi2005abm, iimura2004zfg, knutsson2004pps} -- have been done to make games on peer-to-peer networks more scalable. For example, to reduce traffic between the peers, each peer can send state updates only to whom the state changes are relevant. To achieve this goal, \cite{schiele2007rpp} suggests the division of the virtual environment into regions. The purpose of this is for each region to act as an independent set of smaller scale. Players whose avatars were in a given region could form a small peer-to-peer group and their own computers would decide the outcome of the players' actions in that area of the virtual environment. When an avatar moved from one region to another, it would leave the first group and join the group of its new region. Since the environment is considered contiguous, i.e. the avatars could move freely across adjacent regions, some of the peers in a group would keep connections to peers from groups of neighboring regions and, through these connections between adjacent groups, the necessary information for transferring a player could be obtained.

%TODO: por mais refs aqui após o ", como"
Still according to the work of \cite{schiele2007rpp}, each region has a coordinator. The coordinator is elected among the peers which were present in the region and becomes responsible, only, to decide to whom each state update is relevant -- it does \emph{not} intermediate message exchanges between peers, whose communication is direct. However, this approach is based on the fact that the coordinator is reliable, which can not be guaranteed, since the software used by the player may have been modified to behave incorrectly in order to provide invalid benefits -- what is also known as \emph{cheating}. Another approach would be to elect multiple coordinators per region, but that would involve the implementation of a voting mechanism, and depend on the availability of a certain number of peers to manage the region. Moreover, the communication load remains quadratic (in the worst case), since every peer must send its state to several other peers.

There are also proposals for \textbf{hybrid architectures} \cite{vilanova2008psm, chen2006pch}, using servers and peer-to-peer networks at the same time. According to these works, peers and server share the simulation of the game world. In \cite{vilanova2008psm}, the game world is divided into social space and action spaces. The first is aimed at social interactions, such as chat, exchange of in-game items, forming groups, etc., which are actions that do not impose a heavy burden on the server. However, if the players want, for example, fight -- what is the main objective in most MMOGs -- they have to request to the server the creation of an action space within which the interaction is faster and with little tolerance for communication delays. This space is an \emph{instance} of the game, isolated from the rest of the game world, within which a limited number of players form a small peer-to-peer group, being themselves responsible for performing the simulation and for updating his companions on the game state. To resolve inconsistencies in the simulation due to lack of message ordering, one player is chosen as a \emph{super-peer} and is responsible for ordering the events that are "heavily coupled", whose order can significantly interfere in the outcome of the game. This approach presents some problems, such as the fact that there cannot be a fight, for example, between a large number of players. Moreover, nothing guarantees that the peer chosen to be the super-peer is trustable. If it belongs to a dishonest player, it may put the events in an order that provides an invalid advantage over the other players. In addition, the problem of overloading the bandwidth of peers is addressed by simply setting a limit to the number of participants in each action group, when in fact it is desirable that in an MMOGs there is a lot of players interacting with each other, not only in social circumstances, but also in more dynamic situations, like fast-paced fights.

In \cite{chen2006pch}, the virtual environment is divided into regions, each managed by a peer, which acts as a sub-server. Each one of the other participants in the region sends state updates to this sub-server, which forwards the messages to the other peers in the region to which the state changes may be relevant. To provide fault tolerance, the authors suggest the use of peer clusters, consisting of the region's sub-server and backup peers. These backup peers receive the players' actions from the region manager and process them, creating a backup of the game state. In addition to providing robustness to the system in case the manager crashes, these backup nodes may also detect faults in the simulation and perform recovery procedures. However, this approach does not solve the problem of the peer elected to be the region manager 	not being reliable: to prevent the backup nodes from detecting an invalid simulation, the region manager may send invalid actions to them, instead of forwarding the players' real actions. Finally, if a region manager gets overloaded, it may simply return to the server part of the load assigned to it, keeping a dependence on a centralized infrastructure.

Another hybrid architecture is that of FreeMMG \cite{cecin2004fsa}. In it, the players are organized in a peer-to-peer manner in every region of the virtual environment and the server mediates the communication between different regions. Again, there is the problem of security: as the peers within a region control the simulation that occurs  there, they can also subvert it. The author proposes a probabilistic approach, using a randomly selected peer to verify the simulation in that region. It is expected that, if the peers in a given region try to subvert the game there, at least the randomly inserted node detects the invalid actions and reports to the server. However, besides not completely guaranteeing security, persists the problem that there may be many peers communicating with many, which may compromise the quality of the game.

Some models that suggest the use of clients connected to a distributed server have been proposed in the literature. In \cite{assiotis2006dam}, for example, it is proposed to divide the virtual world into smaller regions, each assigned to a different node of the server system. The authors address the consistency problem by using a lock mechanism -- each entity in the game is associated with a lock that must be obtained before any of the servers making any changes. In addition, they tried to address the problem of \emph{hotspots} (clusters of players in a relatively small area of the virtual environment) through the recursive partitioning of the overpopulated area until the overload is eliminated or there are no more servers available. However, there is a practical limit to this partitioning, for very small areas cause avatars to migrate between regions more often, which implies more network traffic between servers.

Finally, the authors of \cite{assiotis2006dam}, as well as several others with similar proposals \cite{ng2002msa, chertov:olb, lee2003sdl}, consider that the nodes are servers connected through a high speed and low latency network, what makes their solutions only partially applicable in a scenario with highly dynamic and volatile resources, such as the peer-to-peer networks formed with volunteer resources. Such networks have some inherent problems: nodes with low availability and dependability, low bandwidth and low processing power compared to current dedicated MMOG servers maintained by large game companies.

\section{Research questions}
\label{researchquestions}

In a completely decentralized MMOG deployed on a peer-to-peer network, each peer is responsible for deciding the outcome of the actions of its player. Although this makes the system less dependent on a single point of failure -- if a player is disconnected from the game, only the state regarding its avatar will be lost -- and to make the game independent of central servers, this approach proved to be very vulnerable to cheating precisely because at least part of the simulation is relied on the peers. Besides, this approach is prone to saturation of the upload bandwidth of the participants, who must send state updates to their fellow players.

This doctorate plan considers the use of a server system consisting of geographically distributed volunteer nodes. This system would be heterogeneous, where each node could be anything from a home computer to a cluster belonging to some company that had interest in serving the game. Some minimum processing power and communication bandwidth would be required for being part of the server, and it would be necessary to ensure somehow that each node is trusted. However, taking into account that the number of server nodes would be much smaller than the number of players, it would be more feasible to ensure its reliability regarding the possibility of cheating than in a fully decentralized peer-to-peer architecture. Furthermore, the server nodes do not necessarily belong to players. They could be small local Internet service providers (ISPs), for example, who would profit by putting their resources available for a game. That would be interesting to these companies for the resources -- at least communication bandwidth -- of the ISP would be used anyway, for all the traffic from players who were also its subscribers would pass over there regardless of which server each player was connected to.

Nevertheless, as the simulation would be distributed among the server nodes, it is not necessary that all they have the same processing power or communication bandwidth owned by the huge central servers commonly used for MMOGs. To perform this distribution, the game world could be divided into regions, each managed by at least one of these server nodes. More than one server node should be assigned to each region, so that the simulation would be distributed on two levels: first into regions, then within each one of them among the nodes managing it. Thus, each player would send his actions to one of these nodes, which would process them, then synchronize with other nodes in the same region and, finally, make the resulting game state persistent in a distributed way and send it to the players to whom it is relevant. Thus, it would also be provided a better availability, as each region of the virtual world would be served by a set of servers holding a synchronized game state. If any of these servers fail, one of the others could take on the players who were connected to the server that is no longer available.

To achieve the goal of distributing an MMOG server on a dynamic network, with few and volatile resources and less dependability, as are the peer-to-peer networks formed by volunteer nodes, some questions that either have been the subject of research in recent years or that are still unresolved arise. Some of them are:

\begin{itemize}

\item Assign to each server node a portion of the virtual world, so that the load is dynamically balanced, taking into account the existence of \emph{hotspots} (clusters of avatars in a relatively small area);

\item Maintain the simulation consistent among the different regions and among servers managing the same region;

\item Coordinate the entry, exit and the crash detection of nodes of the system, providing fault tolerance;

\item Distribute and store in a persistent manner the state of the world and of the players, so that they can be retrieved when necessary;

\item Optimize the communication between clients and servers in order to save the maximum possible bandwidth, without this being perceptible by the players.

\end{itemize}

It is intended, during the doctoral course, to research and develop new solutions to these problems, seeking those that are most appropriate in a context of massively multiplayer online games. With these issues resolved, it may be provided to MMOGs availability and fault tolerance -- due to the replication of server nodes in the same region -- and scalability, while maintaining the communication delay within an acceptable range for the game, all this for a low cost for the player, compared to traditional MMOG servers.

These solutions will be wrapped together in a middleware, which will provide them as services. These services will be accessible via an API (application programming interface) to allow an easier development of a massively multiplayer game. The game programmer will be able to use these services without needing to have a deep understanding of the distribution model used by the middleware. To the application developer, a high-level interface will be provided, while the middleware will take care of the tasks performed in lower levels. The communication abstraction provided to the game developer may be: publish-subscribe, as in PADRES \cite{fidler2005padres}, distributed shared memory, as in Sinfonia \cite{aguilera2007sinfonia}, message passing or any other that best suits the development of an MMOG.

%\begin{figure}
% \centering
% \includegraphics[width=0.50\textwidth]{imagens/www}
% \caption{Troca de ligações no algoritmo WWW. Esquerda: configuração inicial; Direita: configuração após a troca}
% \label{fig:WWW}
%\end{figure}
%\begin{equation}
%	\label{equ:Hib}
%	 Hib=\frac{1}{N}\sum_{i=1}^NC_i(N_i-N_{0_i})^2 
%\end{equation}

%\begin{equation}
%	\label{equ:Ene}
%	 Energ=\sum_{r_{ij}<r_{max}}[k_r(r_{ij}-r_{0_{ij}})^2+\sum_{r_{ik}<r_{max}}K_a(\theta_{ijk}-\theta_{0_i})^2] 
%\end{equation}

%explicar onde o CoSMMuS se encaixa na literatura: o que que vai contribuir?
%enfiar tantas referências quanto for possível, porém evitando dizer que já existe algo pronto

\chapter{Objective}

The main objective of this doctoral plan is to create a decentralized architectural model to support massively multiplayer online games. Following this model, it should be possible to form a system of geographically distributed server nodes, using the Internet as a means of connection between them. This distribution should be transparent to the players, to whom the system will be seen as a single MMOG server, which maintains the game world consistent, available and with a good quality of service. As a final product, a middleware for MMOG development with a high-level interface will be created.

%Além de distribuir as cargas computacionais (comunicação e processamento) impostas pelo grande número de pessoas jogando, este modelo deverá incluir técnicas com o fim de reduzir o tráfego entre os componentes deste sistema (nodos servidores e máquinas clientes dos jogadores), economizando largura de banda. Tais técnicas são o foco deste trabalho, tendo em vista que se considera um cenário com recursos voláteis e limitados, já que a ligação entre os nodos será feita através da Internet.

%Existem outros aspectos relevantes, além da distribuição do servidor e da economia de largura de banda, que poderão ser acrescentados ao modelo, dependo da disponibilidade de tempo para realizar tal tarefa.

\section{Detailed objectives}
%TODO: por aqui as referências de gerenciamento de interesse e de topologia, difusao etc. (ver slide "em andamento" da semac)

%A função deste middleware será de abstrair para a camada do jogo, que ficará acima dele, o sistema geograficamente distribuído servidor. distribuir as cargas de processamento e comunicação entre 
%
%sistema distribuído sobre a Internet, que funcionará como um grande servidor para jogos online maciçamente multijogador. Pode ser utilizado o conceito de "super-nodo", que nada mais é que um computador conectado à Internet, com capacidade de processamento e comunicação acima da média de usuários comuns, mas que, ainda assim, é bem mais limitado que os grandes servidores que costumam servir os jogos MMOG mais conhecidos. A idéia é que diversos destes super-nodos, conectados à Internet, formem um sistema distribuído que aceitará e repartirá conexões oriundas de jogadores. Eles então dividirão a carga de processamento e de comunicação dos diferentes jogadores conectados, assim como a simulação do ambiente virtual do jogo.

To achieve this general objective, some goals were established:

@@

%TODO: mecanismo -> outra palavra
%TODO: dividir este item em 2: distribuiçao e rede overlay parecida com a real.
\item Develop a mechanism to deal with entry, exit and crash of server nodes, providing fault tolerance, while trying to keep the nodes which are geographically close to each other adjacent in the overlay network of the server system;

%TODO: dividir este item em 2
\item Create an algorithm for balancing the load of the game, maintaining the load on each server proportional to its capacity, while avoiding the splitting of a hotspot into different regions, as this would cause a strong dependence between these regions \cite{bezerra2009lbs};

\item Define a scheme for synchronizing the simulation state among different server nodes assigned to the same region of the gaming world, as well as synchronize periodically, or when necessary, the states in different regions, in order to maintain global consistency;

\item Build a distributed persistence mechanism to store the states of the players and of the virtual game world as a whole, so they are always available for any server node, even if some of these become unexpectedly unavailable;

\item Design techniques and algorithms that aim to reduce intra-server and server-client communication, in order to save resources of the server nodes, reduce delays in the communication between them and expand the audience of the game due to the reduction in the bandwidth required to play it and, besides, reducing the requirements for a computer to be part of the server system, without undermining the game experience of the players;

\item Create a middleware containing the desired services, along with a high-level API to make them accessible to the game developer.

%\item Analisar a aplicabilidade destas técnicas em cenários em que cada nodo do sistema tem recursos mais limitados;
%
%\item Propor integração das mesmas, com otimizações focadas no contexto definido (disponibilidade de super-nodos, ao invés de servidores poderosos);
%
%\item Implementar um protótipo que integre as técnicas estudadas e/ou desenvolvidas no trabalho;
%
%\item Efetuar simulações que demonstrem o resultado obtido do trabalho e avaliar seus resultados.

@@

\chapter{Preliminary model: Cosmmus}
% as refs vão aqui
% divisão em regiões
% distribuição em dois níveis -> porque usar mais de um servidor seria melhor do que um só (menor tráfego entre eles e processamento distribuído)
% Cooperatie System for Massive Multiplayer Service
% duong ta
% citar o fábio e o modelo de sincronização que ele usa (simulador otimista e conservador)
% falar algo de pastry, chord etc., para persistência distribuída
% falar do modelo "filosófico" anti-trapaça (falar das duas opções: provider-centric e anarchy-centric)
% Como os jogadores interagem!!!
% otimização do uso de largura de banda com IM (citar dewan, eu, etc.)

In a previous work \cite{bezerra2009lre}, it was set up a preliminary model -- which aimed to partition the virtual environment into regions, each assigned to a single server -- on which a load balancing scheme was defined. The Cosmmus -- Cooperative System for Massive Multiplayer Service -- is a new model to be defined during the doctorate course, which will also uses the principle of partitioning the virtual environment of the game into regions, but addressing the most critical problems (as referred in section \ref{researchquestions}), which were left out so far.

>>>>

% a data deve ser a da defesa; se nao especificada, são gerados
% mes e ano correntes
\date{September}{2009}

@@

%
% palavras-chave
% iniciar todas com letras minúsculas, exceto no caso de abreviaturas
%
\keyword{massively multiplayer online games}
\keyword{MMOG}
\keyword{distributed server}
\keyword{distributed interactive simulation}

@@

%\begin{abstract}

% This doctorate study plan proposes the research and creation of a novel support model for MMOGs (massively multiplayer online games), by using large-scale systems similar to those created for sharing of computing resources, such as file sharing peer-to-peer networks. Following this model, a distributed system could be formed to act as a server for an MMOG. Then, a portion of the game's communication and processing load would be delegated to each node of the server system -- more precisely, the game's virtual environment would be partitioned into regions, each of which managed by one or more server nodes. Such a system would be perceived by the players as an unique central servel to which they connect via a gateway. Upon connected, each player is seamlessly redirected to the appropriate server node. However, for such server system to work as expected, some issues must be addressed, such as maintaining the game state consistent among the different game server nodes, so each player has the same vision; management and balancing of the load assigned to the different server nodes; what to do when a node crashes or leaves the server system and, finally, how to broadcast and store the game state to the players in a distributed manner so that its most recent state can be restored even in the presence of system failures.

% \end{abstract}

%\chapter{Introduction}

@@

%=============================================================================
%PARÁGRAFO INICIAL
%=============================================================================
\chapter*{Study plan}

This study plan proposes a research project focused on the theme of massively multiplayer online games (MMOGs). Although this kind of games is usually deployed on a client-server infrastructure, with a large and expensive central server, it is possible to use a network with a decentralized architecture. Therefore, ways of providing such decentralized support should be investigated, reducing the maintenance cost of these games, while still providing the basic requirements of MMOGs.

%=============================================================================
%CONTEXTO
%=============================================================================
\section*{Context}

Some years ago, with the broader access to the Internet, the traditional computer games have evolved to multiplayer online games, but with no much more than ten or twenty players in a single match. After that, in the past few years, with faster, cheaper -- when compared to a decade ago --, and more reliable Internet connections, a new game genre has arised as a new trend, which is the MMOG, with potentially tens of thousans of simultaneous players. Some studies demonstrate that the market share of these massively multiplayer games will keep growing at a considerable rate -- according to \cite{intelligence2006online}, the online games market will grow from US\$ 3.4 billion in 2005 to US\$ 13 billion in 2011.

The main reason for the success of these games is the kind of interaction they allow. Usually, such games contain a virtual environment, where its numerous players may interact in many different ways, from fighting -- cooperatively, against artificial intelligence controlled monsters, or even dueling, and fighting each other -- to trading in-game objects -- like weapons and armor in medieval fantasy games -- or just chatting in some place of the game virtual world. This game world, besides, is persistent, i.e. the changes resulting from the player actions will persist, no matter if they keep playing continuously or if they disconnect from the game for a while.

This persitence constitutes one of the main features of these games, whose matches usually do not have any predetermined time to end. There are some exceptions, such as MMORTS (real-time strategy MMOGs), with matches lasting from six months to one year \cite{travian}, what is still considerably long for a game. Most MMOGs, however, have no such time to finish, creating the illusion of a parallel world into which the players may transport themselves when they so desire.

Each player interacts with the virtual environment -- or game world -- by controlling an entity called \emph{avatar}. Each avatar is a representation of a player in the virtual environment -- e.g. a sniper in a war game, a medieval knight in a fantasy game, a gangster or any other controllable character. The avatar performs the actions requested by the player, interfering in the game world and potentially changing its state and the state of other avatars. In the traditional central server model, each player sends his desired actions to the server, which processes them, calculating their outcome, represented as new states for the world and for some avatars. These new states are then broadcast to other players, so they can perceive the changes. The server also acts as an arbiter, who decides the actions performed by non-playable characters -- such as dungeon inhabitants, who might attack anyone who enters there --, whose state is also sent to the players.

It is easy to observe that the central server, receiving actions from thousands of players and broadcasting the resulting state of each of these actions to, potentially, every player, is prone to have its processing power and, specially, its bandwidth saturated by the game. The machine serving the game is, usually, a cluster consisting of nodes connected via a fast and low-latency local network, providing the necessary processing power, and with a very fast Internet connection \cite{feng2007wnn}, whose transmission rate ranges from hundreds of MBps to several GBps, or more. Obviously, such infrastructure is highly expensive, preventing small game developing companies or independent groups with low financial resources to enter the MMOG market.

The purpose of this study plan is, therefore, to develop an innovative solution with a decentralized approach. More specifically, it is intended to create a middleware which uses a decentralized overlay network such as those used for resource sharing -- e.g. Pastry \cite{rowstron2001psd}, Tapestry \cite{zhao2004trg}, Koord \cite{kaashoek2003ksd} and others -- to create a geographically distributed system composed of low-cost volunteer nodes. This system will then act as a geographically distributed server to which the players connect via a gateway. Obviously, this model brings several questions, which were trivial to solve when using the traditional server architecture, but that become much more complex in a geographically distributed server scenario, such as: state consistency among the server nodes; coordination of entry, exit and crash of some of them, in order to keep the service available as much as possible; distributed persistence; optimal management of the scarce resources of the server system; and so on.

%=============================================================================
%POR QUE A SUÍÇA.
%=============================================================================
\section*{Studying in Switzerland}

Dr. Fernando Pedone, associate professor at USI (Universittà della Svizzera italiana), in Lugano, heads a research group working on the area of distributed systems and distributed data management systems. His work has addressed issues related to reliability, as well as replication in such systems by using group communication abstractions. These research topics are closely related to the MMOG study plan outlined here, for the intended MMOG server system has reliability as one of its critical questions to solve. Also, the distributed state persistence and consistency fit on the definition of replicated distributed data management system, only adding some requirements specific to MMOGs (such as timeliness on the retrieval and storage of parts of the game state).

 As another reason for studying in Switzerland, it may be mentioned that the close interaction with researchers from the University of Lugano, working on a closely related area, will provide a good environment to the conception of new ideas regarding the research project presented here. Besides, Prof. Pedone has accepted to work with me in this project and, after several personal meetings and discussions via e-mail, many details of the research plan have been already been discussed, and the granting of a scholarship would be a final need to execute this project.

%=============================================================================
%CRONOGRAMA
%=============================================================================
\section*{Work schedule}

To achieve the goals outlined for the research project, a schedule has been designed, considering a nine-month academic year. This period has been chosen to fit the length of the scholarship to which this application refers. %In Switzerland, it is intended to execute some of the essencial parts of the research project designed for a Ph.D. course in UFRGS (Brazil). This project will, therefore, result in an international cooperation between both UFRGS and the Swiss university.
The tasks to be performed in Switzerland are:

@@

  \item Detailed definition of the services to be provided by the MMOG decentralized support middleware, such as:

    \begin{enumerate}     
     \item Synchronization of the virtual environment simulation among the server nodes;
     \item State storage and retrieval;
     \item Management of the entrance and exit of nodes in the server system;
     \item Fault-tolerance, by allowing the system to keep working correctly even upon unexpected disconnection or crash of some servers;
    \end{enumerate}

  \item Research and development of innovative solutions to deliver these services, considering the needs of MMOGs:

    \begin{enumerate}
      \item Develop algorithms or techniques to manage the consistency of the game world simulation among the server nodes of the system;

      \item Build a distribute persitence scheme, to store and retrieve the persistent state of the game, using replication and providing reliability, while minimizing the delay of each operation;

      \item Design algorithms to reduce the inter-server communication overhead, preventing state-related network messages from overloading the server system;

      \item Create an abstraction for these services, so that the game developer who uses the intended middleware will not have to consider questions regarding the distribution of the server;
    \end{enumerate}

  \item Create a prototype which implements these services;

  \item Validate the proposed middleware by simulations.

@@

Also, besides attending to a local language course, publications are expected. The work schedule for the period in Switzerland is intended to be as follows:

\begin{table}[h]
\begin{center}
%\caption{Créditos a obter}
\label{tab:creditos}
\begin{tabular}{|c||c|c|c|c|c|c|c|c|c|}
\hline
Task   & Sep & Oct & Nov & Dec & Jan & Fev & Mar & Apr & May \\ \hline \hline
Language course & X & X & X &   &   &   &   &   &  \\ \hline
1       & X &   &   &   &   &   &   &   &   \\ \hline
2 (a)   &   & X & X &   &   &   &   &   &   \\ \hline
2 (b)   &   &   &   & X & X &   &   &   &   \\ \hline
2 (c and d) & & &   &   &   & X & X &   &   \\ \hline
3           & & &   &   &   & X & X & X &   \\ \hline
4           & & &   &   &   &   &   &   & X \\ \hline
Writing of papers     & & & X & X & X & X & X & X & X \\ \hline

\end{tabular}
\end{center}
\end{table}

%todas as etapas irão gerar produções bibliográficas

%=============================================================================
%BIBLIOGRAFIA

\bibliography{planodoc}
\bibliographystyle{unsrt}
